Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-27 manuscript/manuscript.Rmd
First pass at edits

Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
Novel data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].
RL methods also bring many challenges of their own: being notoriously difficult to train and evaluate, immense computational costs, and frequent challenges with transfer-ability or reproducibility. 
A review of all these issues is beyond our scope but can be found elsewhere [@rl_intro; @millie_AM-RL].
Here, though, we will focus on the issue of opacity and interpretability raised by these methods.
In contrast with optimization algorithms currently used in either ecosystem management or resource economics, RL algorithms have no guarantees of or metrics for convergence to an optimal solution.
We can only assess the performance of these black box methods relative to alternatives.

@@

Reinforcement learning (RL) is, in a nutshell, a way of approaching *control problems* through machine learning.
An RL algorithm can be conceptually separated into two parts: an *agent*, and an *environment* which the agent can interact with.
That is, the agent may act on the environment and thus change its state, while the environment gives a *reward* to the agent in return (see Fig. \ref{tbd}).
The rewards encode the agent's goal.
The main part of an RL algorithm is then to progressively improve the agent's *policy*, in order to maximize the cumulative reward received. 
This is done by aggregating experience and learning from it.

@@

Mathematically, RL may be formulated using a discrete time *partially observable Markov decision process (POMDP)*.
This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
For the sake of clarity, we will present here only the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
An FMDP may be defined by the following data:

@@

With any fixed policy function, the agent will traverse a path 
$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
sampled randomly from the distribution
\begin{align*}
  p_\pi(\tau) = 
  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
\end{align*}
Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
$$  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
$$
where,
$$
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
$$
is the cumulative reward of path $\tau$.
The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *value function*.

@@

The policy function $\pi$ often lives in a high- or even infinite-dimensional space.
This makes it unfeasible to directly optimize $\pi$.
In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parametrized family of functions.
Deep reinforcement learning uses this strategy, focusing on function families parametrized by neural networks. (See Fig. \ref{tbd}.)
<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->

@@

Reinforcement learning is, on the other hand, typically a model-free approach to control theory.\footnote{There are, however, approaches to perform model-based reinforcement learning. While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
While the model is certainly used to generate training data, it is not directly used by the optimization algorithm.
This provides more flexibility to use model-free RL in instances where the model of the system is not accurately known.
In fact, it has been shown to generally be preferable to use model-free RL in such instances.

@@

Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
This provides very interpretable solutions, such as constant mortality policies.
There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviate considerably from reality.
Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks [@tbd].

We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
Furthermore, rather than computing the optimal control policy for the model (something that is impossible in practice for complex systems) we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.
This policy is estimated in a *model-free* setting, i.e., the agent treats the dynamical model (e.g. eq. \eqref{eq:3d model stoch}) as a black box of input-output pairs.
By not relying on precise knowledge of the model's parameter values, but rather just relying on input-output statistics, model-free approaches have gained traction in a variety of control theory settings [@rl_people].

We compare deep reinforcement learning-based policies against optimal constant mortality policies. 
While the latter are MSY-inspired, they are optimized in a model-free way as well: the optimal constant mortality rate is computed empirically from simulated data.

Because of the simplicity of the constant mortality ansatz, the optimal such policy may be easily estimated (see Figure \ref{fig:tuning}).
In contrast, DRL optimizes over the more expressive -- and more complicated -- family of policies parametrized by a neural network.
Because of this, our DRL approach uses a variant of *gradient ascent* (the proximal policy optimization algorithm) to obtain a good candidate policy.

We have shown that the DRL agent finds an approximately *bang-bang* type of solution (see Textbox 1 for the definition of bang-bang solutions and Figs. \ref{fig:action-X} and \ref{fig:state-space} for the result).
While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.

@@

This manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.

In this regard, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
Importantly, DRL is equipped with tools to deal with complex models for which simple policies might yield unsatisfactory results (and thus more refined techniques are needed to control them).
For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai].

Finally, in our study case, we have shown that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- may be important to maintain the sustainability of the fishery and to increase its long-term profits.
This hypothesis could be explored in more generality as it might apply to a wide variety of models with different optimal control strategies.
In App. \ref{app:tbd} we provide some elementary support for this hypothesis by examining the performance of several ‘‘rule of thumb’’ policies based on conservative corrections to MSY.

>>>>

Recent advances in machine learning may now once again bridge these approaches, while also bringing new challenges of their own. 
Advances in data-driven methods have allowed these models to evolve into ever more complex and realistic simulations used in fisheries management, where models with over 100 parameters are not uncommon [@RamLegacyDB].
Constrained by computational limits, MDP approaches have been intractable on suitably realistic models and largely confined to more academic applications [@Costello2016].
However, advances _Deep Reinforcement Learning_, (RL) a sub-field of machine learning, have recently demonstrated remarkable performance in a range of such MDP problems, from video games [@atari] to fusion reactions [@fusion] to the remarkable dialog abilities of ChatGPT [@chatgpt].

@@

Reinforcement learning (RL) is a way of approaching *sequential decision problems* through machine learning.
All applications of RL can be conceptually separated into two parts: an *agent* and an *environment* which the agent interacts with.
That is, the agent takes actions within the environment.
As a result of an agent's action, the environment will transition to a new state and return a numerical *reward* to the agent (see Fig. \ref{tbd}).
The rewards encode the agent's goal.
The main task of any RL algorithm is then to maximize the cumulative reward received. 
This objective is achieved by aggregating experience in what is called the *training* period and learning from such experience.
<!-- ML: We could cite our deep RL paper here instead of going into further details. Not sure what we are expecting as the audience. -->.

@@

The RL environment can be formally described as a discrete time *partially observable Markov decision process (POMDP)*.
This formalization is rather flexible and allows one, e.g., to account for situations where the agent may not fully observe the environment state, or where the only observations available to the agent are certain functions of the underlying state.
For the sake of clarity, we will only present here the class of POMDPs which are relevant to our work: *fully observable MDPs with a trivial emission function* (FMDPs for short).
An FMDP may be defined by the following data:

@@

With any fixed policy function, the agent will traverse a path 
$\tau=(s_0,\ a_0,\ s_1,\ a_1\ \dots,\ s_{t_{\text{fin.}}})$ 
sampled randomly from the distribution
\begin{align*}
  p_\pi(\tau) = 
  d(s_0) \prod_{t=0}^{ t_{\text{fin.}}-1 }
  \pi( a_t | s_t ) T( s_{t+1} | s_t, a_t ).
\end{align*}
Reinforcement learning seeks to optimize $\pi$ such that the expected rewards are maximal,
$$  
  \pi^* = \mathrm{argmax}\ \mathbb{E}_{\tau\sim p_\pi}[R(\tau)],
$$
where,
$$
  R(\tau) = \sum_{t=0}^{ t_{\text{fin.}}-1 } \gamma^t r(a_t, s_t),
$$
is the cumulative reward of path $\tau$.
The function $J(\pi):=\mathbb{E}_{\tau\sim p_\pi}[R(\tau)]$ is called the *expected return*.

@@

The policy function $\pi$ often lives in a high or even infinite-dimensional space.
This makes it unfeasible to directly optimize $\pi$.
In practice, an alternative approach is used: $\pi$ is optimized over a much lower-dimensional parameterized family of functions. <!-- ML: I am not following this line of thought here. I see the main reason why NN's are used as being that NN's are universal function approximators, and we don't know what the policy and/or value function will look like.-->
Deep reinforcement learning uses this strategy, focusing on function families parameterized by neural networks. (See Fig. \ref{tbd}.)
<!-- The figure should kinda explain how the parametrization arises, and how gradient ascent is used. Maybe namedrop backpropagataion for the sake of usefulness to the reader? -->

@@

Reinforcement learning, on the other hand, can provide a model-free approach to control theory.\footnote{There are approaches to perform model-based reinforcement learning. While we will not focus on these in this paper they are discussed in \cite{tbd1, tbd2}.}
While a model is often used to generate training data, this model is not directly used by model-free RL algorithms.
This provides more flexibility to use RL in instances where the model of the system is not accurately known.
In fact, it has been shown that model-free RL outperforms model-based alternatives in such instances \cite{Janner et al., 2019}.

@@

Fisheries are complex ecosystems, with interactions between species leading to highly non-linear dynamics.
While current models for population estimation include many important aspects of this complexity, it is still common to use simplified dynamical models in order to guide decision making.
This provides very interpretable solutions, such as constant mortality policies.
There is a drawback here, however: due to the simplicity of these dynamical models, the policies might not respond effectively in many situations---situations where the predictions of these simple models deviates considerably from reality.
Because of this, policies such as MSY have faced pushback and are believed to have contributed to the depletion of fish stocks.

We propose an alternative approach to the problem of fishery control: to use a more expressive -- albeit more complicated -- dynamical model to guide decision making.
Furthermore, rather than computing the optimal control policy for the model -- something that is impossible in practice for complex systems -- we use deep reinforcement learning to obtain a ‘‘pretty darn good’’ policy.

In this paper we have exemplified this approach by applying it to the control of a three-species fishery model.
In this example, we have shown that the DRL agent finds an approximately *bang-bang* type of solution.
While bang-bang solutions are known to be optimal for a wide variety of single-species fishery control problems [@alexandru], there is no such proof of optimality for general multi-species models.
In this sense this example is rather surprising and reassuring: while no general guarantee points at the optimality of bang-bang solutions for multi-species model, the existence of one such example provides a benchmark for the use of DRL in fishery control.

@@

This way, this manuscript functions as a first step in the process of incorporating DRL into fishery decision making.
Importantly, DRL is equipped with tools to deal with more complicated models which might require a more refined approach to control them.
For example *long short-term memories (LSTMs)* may be used to incorporate not only $N_t$ but also $\mathrm{d}N_t/\mathrm{d}t$ into policy decisions---while this additional degree of freedom would make analytical control theory approaches unwieldy, it may be easily and cost-effectively implemented in DRL.
Similarly, *curriculum learning* could be incorporated to deal with phenomena such as parametric noise or parametric drift [@openai]. <!-- ML: Could also use LSTMs for parametric drift. Might be better to use a different selling point for curriculum learning like if you want to have a bit more generalizability in your RL agent. -->

This way, DRL provides a framework flexible enough to capture the intricate population dynamics that are relevant in fishery management.
We have shown that highly non-linear systems may give rise to simplified and good policies and that these may be found via reinforcement learning.
Moreover, the framework allows for the possibility that these simplifications do not in fact happen, and that good harvest policies are highly non-linear, multi-parameter functions.

Conversely, our analysis suggests that, for high dimensional systems, bang-bang policies lead to increased sustainability and long-term profits as compared to constant mortality strategies.
Similarly, in App. \ref{app:tbd} we investigate several generalizations of the constant effort approach. 
For example, we examine strategies with a ramp-down from constant effort to zero effort at low populations.
These, too, yield subpar results as compared to escapement.
This seems to indicate that the ‘‘rest periods’’ so prominent in both the DRL solution and in the optimal bang-bang policy -- periods of several time steps with no fishing at all -- are important to maintain the sustainability of the fishery and, additionally, increase long-term profits.
